---
title: "Springboard Data Wrangling Exercise 2 - Titanic"
author: "Chinpei Tang"
date: "May 19, 2016"
output: pdf_document
---
## Set working directory
First, save the titanic3.xls file to titanic_original.csv file, and set to the correct working directory.
```{r}
setwd("C:/Users/Chinpei/Documents/GitHub/Springboard_FDS/DW_Ex2")
```

## Load original data 
Since "Titanic" is one of the preloaded dataset in RStudio, and to avoid overriding the dataset, the dataset is imported as "titanic_ex" (_ex means exercise). Also assign blank and space data to NA.
```{r}
titanic_ex = read.csv("titanic_original.csv", header = T, na.strings = c(""," "))
```
Examine the data.
```{r}
dim(titanic_ex)
summary(titanic_ex)
```
There are 1310 observations, and 14 columns.

## Port of embarkation
Examine the NA's in embarked column.
```{r}
summary(titanic_ex$embarked)
```
In fact, there are actually 3 missing values instead of 1. Substitute the missing port of embarkation to "S". 
```{r}
titanic_ex$embarked[is.na(titanic_ex$embarked)] = "S"
summary(titanic_ex$embarked)
```
Now there is no more NA.

## Age
Examine the NA's in age column.
```{r}
summary(titanic_ex$age)
```
There are 264 NA entries. Calculate the mean of the age ignoring the NA's.
```{r}
mean(titanic_ex$age, na.rm = T)
```
Substitute the NA's with the mean values.
```{r}
titanic_ex_agemean = titanic_ex
titanic_ex_agemean$age[is.na(titanic_ex$age)] = mean(titanic_ex$age, na.rm = T)
summary(titanic_ex_agemean$age)
```
Some other ways to populate the missing values are taking the median value.
```{r}
titanic_ex_agemed = titanic_ex
titanic_ex_agemed$age[is.na(titanic_ex$age)] = median(titanic_ex$age, na.rm = T)
summary(titanic_ex_agemed$age)
```
However, there is not much statistical difference between taking median or mean values. So use the mean age to substitute the average data.
```{r}
titanic_ex = titanic_ex_agemean
```

## Lifeboat
NA has already been assigned to the blank or space data in the boat column.
```{r}
summary(titanic_ex$boat)
titanic_ex$boat[is.na(titanic_ex$boat)]
summary(titanic_ex$boat)
```

## Cabin
Finally, create the "has_cabin_number" column for the passenger with cabin numbers.
```{r}
summary(titanic_ex$cabin)
titanic_ex$has_cabin_number = as.integer(!is.na(titanic_ex$cabin))
```

## Write to clean file
Write the new dataset to the clean csv file.
```{r}
write.csv(titanic_ex, file = "titanic_clean.csv")
```